ANGSD-wrapper: utilities for analyzing next generation sequencing data

### 2016-01-17 manuscript.tex
changed to PDFs
draft done

A number of tools have recently been published to handle high throughput sequencing data \citep{garrigan2013popbam, purcell2007plink, danecek2011variant, hutter2006genome}, but the majority of these either make limiting assumptions about the data (e.g., all sites have been sequenced, all genomes are haploid, sequencing is to sufficient depth, all individuals are outcrossing) or are specialized tools offering a narrow set of analysis options. 
Korneliussen et al. \citep{korneliussen2014angsd} recently published the software package ANGSD, which enables users to flexibly perform a large number of common population genetic analyses, including diversity statistics, admixture analysis including Patterson's D statistic \citep{Durand:2011jd}, site frequency spectrum estimation \citep{pmid22911679}, and neutrality test statistics \citep{korneliussen_calculation_2013}. %\jri{are all these within ANGSD or are some ngsPopgen?} all are in ANGSD
One of the most important features of ANGSD is that most analyses are performed directly on genotype likelihoods, freeing users from the requirement of calling variants or genotypes and permitting analysis of low-coverage data or sequences with large amounts of missing data.

@@

\begin{figure}
\centering
\includegraphics[width=130mm]{figures/fig1.png}
\caption{A visualization of Watterson's $\theta$ estimated by ANGSD across a 1.5 megabase region of chromosome 10 in {\it Zea mays} spp. {\it mays} using ANGSD-wrapper. Blue boxes indicate genic regions provided by a GFF annotation. \label{fig:theta} }
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/mt4labeled.png}
\caption{Admixture analysis for {\it Zea mays} spp. {\it mays}, {\it Zea mays} spp. {\it parviglumas}, and {\it Zea mays} spp. {mexicana} (\texit{mex}) with K=4 source populations. For a list of samples see Table \ref{tab:samples}. }
\label{fig:admixture}
\end{figure}

@@

We first use the SFS method to estimate the site frequency spectrum (SFS) of both maize and its wild progenitor \textit{Zea mays} ssp. \textit{parviglumis}, assuming an inbreeding coefficient of $F=1$ for these highly inbred samples. 
Consistent with previous results \citep{hufford2012comparative}, we show that the maize SFS is skewed towards more intermediate-frequency variants (Figure \ref{fig:figure3}A-B), likely a result of the bottleneck associated with maize domestication \citep{Beissinger031666}.
We find further evidence of the effect of domestication using the Thetas method, revealing lower overall levels of diversity in maize (Figure \ref{fig:figure3}C). 
These results use a sample from the genus \textit{Tripsacum} as an ancestral sequence generated using the Ancestral Sequence method. 
Using the 2DSFS method, which includes an \fst calculation, we then scan our example region for windows with elevated differentiation between maize and teosinte. 
The mean \fst in this region is 0.013, which is an order of magnitude lower than the genome-wide reported value of 0.11 \citep{hufford2012comparative}.   
The \X windows in the 1\% tail of highest \fst values include genes several genes identified as potential targets of selection during domestication in \citep{hufford2012comparative}

@@

\section*{Acknowledgements}
We acknowledge funding support from the US National Science Foundation (IOS-1238014 to JRI and IOS-1339393 to PLM) and funding from the UC Davis Plant Sciences Department.  We thank members of the Ross-Ibarra and Morrell labs for discussion and software testing. We thank the authors of ANGSD and related programs for answering questions, particularly Matteo Fumagalli and Filipe Vieira. Finally, we would like to thank Felix Andrews for statistical advice, although we did not follow it. 
\jri{make sure figures/tables go where Mol. Ecol. Res. suggests (at end? in text?)}
\clearpage
\bibliographystyle{unsrt}
\bibliography{references}

>>>>

A number of tools have recently been published to handle high throughput sequencing data \citep{garrigan2013popbam, purcell2007plink, danecek2011variant, hutter2006genome}, but the majority of these either make limiting assumptions about the data (e.g., all sites have been sequenced, all genomes are haploid, sequencing is to sufficient depth, all individuals are outcrossing) or are specialized tools offering a narrow set of analysis options. 
Korneliussen et al. \citep{korneliussen2014angsd} recently published the software package ANGSD, which enables users to flexibly perform a large number of common population genetic analyses, including diversity statistics, admixture analysis including Patterson's D statistic \citep{Durand:2011jd}, site frequency spectrum estimation \citep{pmid22911679}, and neutrality test statistics \citep{korneliussen_calculation_2013}. 
One of the most important features of ANGSD is that most analyses are performed directly on genotype likelihoods, freeing users from the requirement of calling variants or genotypes and permitting analysis of low-coverage data or sequences with large amounts of missing data.

@@

\begin{figure}
\centering
\includegraphics[width=130mm]{figures/fig1.pdf}
\caption{A visualization of Watterson's $\theta$ estimated by ANGSD across a 1.5 megabase region of chromosome 10 in {\it Zea mays} spp. {\it mays} using ANGSD-wrapper. Blue boxes indicate genic regions provided by a GFF annotation. \label{fig:theta} }
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/mt4labeled.pdf}
\caption{Admixture analysis for {\it Zea mays} spp. {\it mays}, {\it Zea mays} spp. {\it parviglumas}, and {\it Zea mays} spp. {mexicana} (\texit{mex}) with K=4 source populations. For a list of samples see Table \ref{tab:samples}. }
\label{fig:admixture}
\end{figure}

@@

We first use the SFS method to estimate the site frequency spectrum (SFS) of both maize and its wild progenitor \textit{Zea mays} ssp. \textit{parviglumis}, assuming an inbreeding coefficient of $F=1$ for these highly inbred samples. 
The SFS and diversity statistics were calculated using ancestral states inferred in the Ancestral Sequence method from a single resequenced genome of \textit{Tripsacum}.  
Consistent with previous results \citep{hufford2012comparative}, we show that the maize SFS is skewed towards more intermediate-frequency variants (Figure \ref{fig:figure3}A-B), likely a result of the bottleneck associated with maize domestication \citep{Beissinger031666}.
We find further evidence of the effect of domestication using the Thetas method, revealing lower overall levels of diversity in maize (Figure \ref{fig:figure3}C). 
Using the 2DSFS method, which includes an \fst calculation, we find a mean \fst in this region of 0.116, nearly identical to the  genome-wide value of 0.11 reported in \citet{hufford2012comparative}.   
There are no loci in this region that have previously been identified as showing evidence of selection during domestication, and these data also do not find any large consecutive regions of high \fst.

@@

\section*{Acknowledgements}
We acknowledge funding support from the US National Science Foundation (IOS-1238014 to JRI and IOS-1339393 to PLM) and funding from the UC Davis Plant Sciences Department.  We thank members of the Ross-Ibarra and Morrell labs for discussion and software testing. We thank the authors of ANGSD and related programs for answering questions, particularly Matteo Fumagalli and Filipe Vieira. Finally, we would like to thank Felix Andrews for statistical advice, although we did not follow it. 
\clearpage
\bibliographystyle{unsrt}
\bibliography{references}
